---
title: "LibeRtAD automatic-differentiation engine"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LibeRtAD automatic-differentiation engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose and architecture

LibeRtAD is the automatic-differentiation layer used by LibeRation. Model
text is translated to a validated, serializable intermediate representation
(IR). C++ then owns the executable program and persistent CppAD tape. R owns a
small R6 wrapper containing external pointers and the IR required to rebuild
those pointers in a worker process.

The separation is intentional:

1. model text and IR can be saved or sent to another process;
2. arbitrary R evaluation is excluded from the numerical path; and
3. values, Jacobians, gradients, and Hessians are evaluated in C++.

## Supported model language

The first compiler version accepts scalar assignments, arithmetic operators,
common mathematical functions, and `ifelse()` conditions. NONMEM-style
indexed parameters are normalized to ordinary IR inputs.

```{r language}
library(LibeRtAD)
ad_supported()

ir <- ad_ir(
  paste(
    "CL = THETA(1) * exp(ETA(1))",
    "V = THETA(2)",
    "K = CL / V",
    "PENALTY = log(K)^2",
    sep = "\n"
  ),
  outputs = c("K", "PENALTY")
)
ir
```

Undeclared scalar symbols become inputs in first-use order. Supplying
`inputs` explicitly is preferable at interfaces where a stable domain order
matters. Unknown function calls, arbitrary R code, loops, allocation, and I/O
are rejected rather than evaluated.

## Compiling and recording a tape

`ad_compile()` creates an `ADModel`. Passing `at` also records a tape. The
`wrt` argument selects the differentiable domain and `outputs` selects the
range.

```{r compile}
model <- ad_compile(
  "CL = THETA(1) * exp(ETA(1))\nPENALTY = log(CL)^2",
  at = c(THETA_1 = 2, ETA_1 = 0),
  wrt = c("THETA_1", "ETA_1"),
  outputs = "PENALTY"
)
model

point <- c(THETA_1 = 2.2, ETA_1 = 0.1)
model$value(point)
model$gradient(point)
model$hessian(point)
model$value_gradient(point)
```

A multi-output tape exposes a Jacobian. A gradient or Hessian requires exactly
one active output.

```{r jacobian}
multi <- ad_compile(
  "SUM = X + Y\nPRODUCT = X * Y",
  inputs = c("X", "Y"),
  outputs = c("SUM", "PRODUCT"),
  at = c(X = 2, Y = 3),
  wrt = c("X", "Y")
)
multi$jacobian(c(X = 4, Y = 5))
```

Use `$record()` to change the differentiation domain or output set without
recompiling the IR.

```{r rerecord}
multi$record(
  at = c(X = 2, Y = 3),
  wrt = "X",
  outputs = "PRODUCT"
)
multi$value_gradient(c(X = 4))
```

## Worker reconstruction and lifetime

External pointers are process-local and must not be serialized. Persist
`model$ir`, or the original model text, and rebuild with `ADModel$new(ir)` in
the destination process. This is the pattern used by LibeRation and
LibeRties workers.

The R6 object controls pointer lifetime. Keep the object alive for as long as
its compiled program or tape is needed. Replacing or removing the object lets
R finalize the C++ resources.

## Engine information

```{r engine-info}
ad_engine_info()
```

The package is intentionally a numerical backend. Population objectives,
ADVAN event handling, estimation algorithms, and graphical workflows belong
to LibeRation.
