---
title: "Getting Started with LibeRtAD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with LibeRtAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

LibeRtAD provides reverse- and forward-mode automatic differentiation for
scalar objectives in R. The main entry point is `autodiff()`; `backdiff()`
and `forwarddiff()` are convenience wrappers.

## Reverse mode

Differentiate a simple function at a point:

```{r reverse-scalar}
library(LibeRtAD)

f <- function(x, y, z) {
  3 * x^4 + y / z / x + z * 3
}

res <- backdiff(f, x = 2, y = 2, z = 3)
res$value
res$gradient
```

Vector arguments are supported; partial derivatives are returned per element:

```{r reverse-vector}
f_vec <- function(x) sum(x^2)
res_vec <- backdiff(f_vec, x = c(1, 2, 3))
res_vec$partials
```

## Forward mode and the C++ backend

Forward mode propagates tangents from each parameter. The C++ backend can
accelerate the backward or forward pass:

```{r forward-cpp}
f_sqrt <- function(x) sum(sqrt(abs(x)))
fwd <- autodiff(
  f_sqrt,
  x = c(1, 4, 9),
  mode = "forward",
  backend = "cpp"
)
fwd$partials
```

## Matrix arguments

Matrix and vector parameters work the same way:

```{r matrix-args}
f_mat <- function(A, B) sum(A %*% B)
A <- matrix(c(1, 2, 3, 4), nrow = 2)
B <- matrix(c(5, 6, 7, 8), nrow = 2)
mat_res <- autodiff(f_mat, A = A, B = B, mode = "reverse")
names(mat_res$partials)
```

## Hessian

For small parameter vectors, `autodiff_hessian()` builds the Hessian from AD:

```{r hessian}
f_quad <- function(x) sum(x^2)
H <- autodiff_hessian(f_quad, x = c(1, 2, 3))
H
```

## Piecewise expressions

Non-smooth functions such as `pmax()` are handled on the active branch:

```{r piecewise}
f_pw <- function(x) pmax(x^2, 2 * x)
pw <- backdiff(f_pw, x = 3)
pw$value
pw$gradient
```

## Tape diagnostics

When `record_tape = TRUE` (the default), you can inspect tape statistics:

```{r tape-stats}
reset_tape()
backdiff(f, x = 2, y = 2, z = 3, record_tape = TRUE)
ad_tape_stats()
```

## Tape reuse

For repeated evaluations with the same graph structure (e.g. optimisation loops),
save and reuse the tape to skip graph construction:

```{r tape-reuse}
f_quad <- function(x) sum(x^2)
autodiff(f_quad, x = c(1, 2), record_tape = TRUE)
ad_tape_save("quad2")
ad_tape_reuse(f_quad, at = list(x = c(3, 4)), cache_key = "quad2")
```

## NONMEM expression syntax

Control streams often use Fortran `**` for powers. Normalize before AD:

```{r nm-expr}
ad_nm_expr_normalize("CL = THETA(1) ** 2 * EXP(ETA(1))")
```

## Sparse Jacobian

For vector-valued residuals, extract sparsity pattern and values:

```{r sparse-jac}
f_res <- function(theta) c(theta[1]^2 - 4, theta[2] - 1)
sparse_jacobian(f_res, theta = c(2, 1))
```

## Next steps

- Function reference: `?autodiff`, `?backdiff`, `?autodiff_hessian`, `?ad_tape_reuse`
- Example script: `system.file("examples", "basic-example.R", package = "LibeRtAD")`
- Linked packages can register custom reverse rules via C callables (see package source)
