---
title: "LibeRtAD automatic-differentiation engine"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LibeRtAD automatic-differentiation engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose and architecture

LibeRtAD is the automatic-differentiation layer used by LibeRation. Model
text is translated to a validated, serializable intermediate representation
(IR). C++ then owns the executable program and persistent CppAD tape. R owns a
small R6 wrapper containing external pointers and the IR required to rebuild
those pointers in a worker process.

The separation is intentional:

1. model text and IR can be saved or sent to another process;
2. arbitrary R evaluation is excluded from the numerical path; and
3. values, Jacobians, gradients, and Hessians are evaluated in C++.

## Supported model language

The first compiler version accepts scalar assignments, arithmetic operators,
common mathematical functions, and `ifelse()` conditions. NONMEM-style
indexed parameters are normalized to ordinary IR inputs.

```{r language}
library(LibeRtAD)
ad_supported()

ir <- ad_ir(
  paste(
    "CL = THETA(1) * exp(ETA(1))",
    "V = THETA(2)",
    "K = CL / V",
    "PENALTY = log(K)^2",
    sep = "\n"
  ),
  outputs = c("K", "PENALTY")
)
ir
```

Undeclared scalar symbols become inputs in first-use order. Supplying
`inputs` explicitly is preferable at interfaces where a stable domain order
matters. Unknown function calls, arbitrary R code, loops, allocation, and I/O
are rejected rather than evaluated.

## Compiling and recording a tape

`ad_compile()` creates an `ADModel`. Passing `at` also records a tape. The
`wrt` argument selects the differentiable domain and `outputs` selects the
range.

```{r compile}
model <- ad_compile(
  "CL = THETA(1) * exp(ETA(1))\nPENALTY = log(CL)^2",
  at = c(THETA_1 = 2, ETA_1 = 0),
  wrt = c("THETA_1", "ETA_1"),
  outputs = "PENALTY"
)
model

point <- c(THETA_1 = 2.2, ETA_1 = 0.1)
model$value(point)
model$gradient(point)
model$hessian(point)
model$value_gradient(point)
```

A multi-output tape exposes a Jacobian. A gradient or Hessian requires exactly
one active output. LibeRtAD selects multi-direction Forward sweeps for suitable
dense problems and subgraph Reverse for sufficiently large sparse Jacobians;
`$tape_info()` reports the strategy used.

```{r jacobian}
multi <- ad_compile(
  "SUM = X + Y\nPRODUCT = X * Y",
  inputs = c("X", "Y"),
  outputs = c("SUM", "PRODUCT"),
  at = c(X = 2, Y = 3),
  wrt = c("X", "Y")
)
multi$jacobian(c(X = 4, Y = 5))
multi$tape_info()
```

## Dynamic parameters

Inputs omitted from `wrt` are recorded as CppAD dynamic parameters. Their
values can change without reconstructing the operation graph, while
derivatives remain defined only with respect to the active domain. This is
used by LibeRation to share structural prediction tapes across subjects whose
event topology is the same but whose numeric covariates differ.

```{r dynamic}
dynamic <- ad_compile(
  "Y = X^P + P^X",
  inputs = c("X", "P"), outputs = "Y",
  at = c(X = 2, P = 0), wrt = "X"
)
dynamic$set_dynamic(c(P = 3))
dynamic$value_gradient(c(X = 2))
```

Use `$record()` to change the differentiation domain or output set without
recompiling the IR.

```{r rerecord}
multi$record(
  at = c(X = 2, Y = 3),
  wrt = "X",
  outputs = "PRODUCT"
)
multi$value_gradient(c(X = 4))
```

## Worker reconstruction and lifetime

External pointers are process-local and must not be serialized. The portable
choices are the model IR or an optimized graph cache. `$save_tape()` stores the
IR, graph, domain/range metadata, dynamic values, and exact bundled-CppAD
version and commit. `ad_load_tape()` validates that provenance before
reconstructing a ready-to-evaluate tape without parsing or retaping.

```{r cache}
path <- tempfile(fileext = ".rds")
dynamic$save_tape(path)
restored <- ad_load_tape(path)
restored$value_gradient(c(X = 2, P = 4))
unlink(path)
```

The R6 object controls pointer lifetime. Keep the object alive for as long as
its compiled program or tape is needed. Replacing or removing the object lets
R finalize the C++ resources.

## Engine information

```{r engine-info}
ad_engine_info()
```

## Benchmarking

`ad_benchmark()` measures recording separately from repeated tape calls. The
built-in workloads cover nonlinear curvature, a PK-shaped scalar objective,
and a multi-output Jacobian. Central finite differences are included as an
independent numerical agreement check and a call-overhead comparator.

```{r benchmark, eval=FALSE}
ad_benchmark_cases()
result <- ad_benchmark("pk", iterations = 1000, warmups = 50)
result

libertad_gui()
```

The purple benchmark workbench also detects the repository's paired
LibeRation/NONMEM harness when it is available. That benchmark is launched in
a separate, cancellable R process and retains its reports and raw results.
Population objectives, ADVAN event handling, and estimation algorithms remain
the responsibility of LibeRation; the GUI is a measurement surface rather
than a second numerical engine.

`ad_checkpoint_probe()` compares nested-AD-safe `chkpoint_two` prototypes with
directly taped ADVAN1 and 2x2 matrix updates. These probes are experimental:
the production path continues to use direct specialized kernels until
representative benchmarks show that checkpointing offsets its call overhead.
